### 2016-09-23 paper.tex
non-asymptotic

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\vecspan}{span}
\DeclareMathOperator*{\affspan}{aff}
\DeclareMathOperator*{\tr}{tr}

\newcommand{\zero}{\text{\textbf{0}}}
\newcommand{\Zproj}{Z^{\textit{proj}}}
\newcommand{\Zreopt}{Z^{\textit{owa}}}
\newcommand{\nreopt}{n^{\textit{owa}}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\W}{{\hat \Theta^{\textit{owa}}}}
\newcommand{\Waff}{\mathcal{\hat A}}
\newcommand{\WaffE}{{\mathcal{\hat A}_\E}}
\newcommand{\Wave}{{\mathcal{\hat W}^{ave}}}
\newcommand{\Wtave}{{\mathcal{W}^{ave,*}}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\theta}
\newcommand{\wkl}{\hat\w^{kl}}
\newcommand{\ahat}{\hat\alpha}
\newcommand{\afull}{\ahat^{\textit{full}}}
\newcommand{\wreopt}{\hat\w^{owa}}
\newcommand{\wreoptfull}{\hat\w^{\textit{owa,full}}}
\newcommand{\wave}{\hat\w^{ave}}
\newcommand{\wtave}{\E\hat\w^{ave}}
\newcommand{\waver}{\hat\w^{ave,r}}
\newcommand{\wboot}{\hat\w^{boot}}
\newcommand{\wmle}{\hat\w^{mle}}
\newcommand{\wmler}{\hat\w^{mle,r}}
\newcommand{\wstar}{{\w^{*}}}
\newcommand{\wq}{\hat\w^{q}}
\newcommand{\wqstar}{\hat\w^{q^*}}
\newcommand{\dist}{\mathcal{D}}

@@

\newcommand{\I}{\mathcal I}
%\newcommand{\I^{-1}}{\I^{-1}}
\newcommand{\law}{\ensuremath{\xrightarrow{L}}}
\newcommand{\normal}[2]{\ensuremath{\mathcal{N}\left({{#1}},{{#2}}\right)}}
\newcommand{\trans}[1]{\ensuremath{{#1}^{\mathsf{T}}}}
\newcommand{\ltwo}[1]{{\left\lVert {#1} \right\rVert}}
\newcommand{\lone}[1]{{\lVert {#1} \rVert}_1}
\newcommand{\lzero}[1]{{\lVert {#1} \rVert}_0}
\newcommand{\proj}[1]{\pi_{{#1}}}
\newcommand{\prob}[1]{\Pr\left[{#1}\right]}

@@

@@

\begin{abstract}
We present a distributed learning algorithm with low communication complexity.
Our algorithm is an easy to implement modification of the standard averaging method.
In the averaging method, each machine in a cluster first solves the learning problem on a subset of the data;
the resulting parameter vectors are then averaged together.
In our method, we perform a second round of optimization to find the \emph{optimal weighted average} of the parameter vectors.
This second optimization happens in a low dimensional space,
and so it adds negligible computational and communication costs.
%Given $m$ machines (each with $n$ data points of dimension $d$),
%we show that our algorithm reduces bias at a rate of $O(\frac{1-m/d}{n})$.
Whereas the averaging method reduces only the variance of the estimator,
our method reduces both the bias and the variance.
Our analysis is more general than the analysis of similar communication efficient algorithms that also reduce the bias.
Notably, we do not require the likelihood function to be convex.
In practice,
this bias reduction makes our algorithm significantly less sensitive to the amount of regularization.
This has the practical benefit of greatly speeding up the model selection process.
\end{abstract}

@@

\emph{Non-interactive} algorithms require only a single round of communication.
They are significantly faster than interactive algorithms
but have worse regret bounds.
Recent work (discussed in Section \ref{sec:bounds}) has shown that no non-interactive algorithm can achieve regret bounds comparable to an interactive one.
The popular large scale learning system Vowpal Wabbit cleverly uses both types of algorithms to achieve good real world performance.
A non-interactive algorithm is run first to provide a good warm start to a subsequent interactive algorithm \citep{vw,agarwal2014reliable}.
%Synonyms for non-interactive algorithms used in the literature include \emph{one-shot distributed learning} and \emph{communication efficient learning}.

In this paper, we propose a \emph{semi-interactive} distributed algorithm called XXX.
Our algorithm performs two rounds of communication,
so it is not subject to the existing regret bounds of non-interactive algorithms.
The amount of information communicated in the second round is small, however,
so our algorithm retains the speed advantages of non-interactive algorithms.

@@

To motivate our estimator, we first describe a simpler estimator that uses the entire dataset for the second round of optimization.
%The idea of our proposed solution is to find the best linear combination of the $\wmle_i$s using a second round of optimization.
Define the matrix $\hat W : \mathbb{R}^{d\times m}$ to have $i$th column equal to $\wmle_i$.
Now consider the estimator
\begin{equation}
\wreoptfull = \hat W \afull
\end{equation}
where
\begin{equation}
\afull = \argmax_\alpha \sum _{(\x,y)\in Z} f\left(y,\trans\x \hat W \alpha \right)
\end{equation}
%This is the maximum likelihood estimator when we restrict our
Notice that $\wreoptfull$ is the maximum likelihood estimator when the parameter space $\Theta$ is restricted $\W = \vecspan \{\wmle_i\}_{i=1}^m$.
Figure \ref{fig:contour} shows graphically that no other estimator in $\W$ can have higher empirical likelihood than $\wreoptfull$.

Calculating the weights $\afull$ directly is infeasible because it requires access to the full dataset.
Fortunately, we don't need to consider all the data points for an accurate estimator.
The parameter space $\W$ is $m$-dimensional.
So intuitively, we only need $O(m)$ data points to solve such a problem to our desired accuracy.
%(This intuition is formalized in Section \ref{sec:anal}.)
This intuition motivates the following estimator.
Let $\Zreopt_i\subset Z_i$ be a set of $\nreopt$ data points uniformly sampled from $Z_i$ without replacement,
and let $\Zreopt$ be the union of the $\Zreopt_i$s.
Then the OWA estimator is defined as
\begin{equation}
\wreopt = \hat W \ahat
\end{equation}
where
\begin{equation}
\ahat = \argmax_\alpha \sum _{(\x,y)\in \Zreopt} f\left(y,\trans\x \hat W \alpha \right)
\end{equation}

Algorithm \ref{fig:alg2} shows the steps for efficiently computing $\wreopt$.
In the first round, each machine calculates $\wmle_i$ independently and broadcasts the results to every other machine.
Since the parameter vector has $d$ dimensions and there are $m$ machines, a total of $O(md)$ bits of get transmitted.
In the second round, each machine projects its local dataset $\Zreopt$ onto the space $\W$.
These projected data points are then transmitted to a predesignated master machine.
The projected data points each have dimension $m$, so a total of $O(m^2\nreopt)$ bits are transmitted.
The master machine now has all of the information to complete the optimization.

@@

\node[dot] at (-1.35,2.03) {};
\node at (-0.9,2.4) {$\wreoptfull$};

@@

\end{tikzpicture}
\caption{
    Our method performs a second round of optimization to find the best parameter vector in $\W$.
    Since $\wave\in\W$, the estimator $\wreoptfull$ has higher empirical likelihood than $\wave$ almost surely.
}
\label{fig:contour}
\end{figure}

\begin{algorithm}[t]
\caption{Estimation in two rounds}
\label{alg:distributed}
\begin{algorithmic}
\State Each machine $i$ independently
\State ~~~~~loads dataset $Z_i$
\State ~~~~~calculates $\wmle_i$
\State All machines broadcast $\wmle_i$ to all other machines
\State Each machine $i$ independently
\State ~~~~~constructs $\hat W$
\State ~~~~~randomly selects a dataset $Z^{reopt}_i\subset Z_i$
\State ~~~~~calculates $\Zproj_i=\{(\trans\x\hat W,y) : (\x,y)\in\Zreopt\}$
\State All machines broadcast $\Zproj_i$ to a master machine
\State The master calculates $\wreopt$
\end{algorithmic}
\label{fig:alg2}
\end{algorithm}

%\begin{algorithm}[t]
%\caption{Estimation with common data}
%\label{alg:distributed}
%\begin{algorithmic}
%\State Each machine $i$ independently
%\State ~~~~~loads dataset $Z_i$
%\State ~~~~~calculates $\wmle_i$
%\State All machines broadcast $\wmle_i$ to all other machines
%\State Each machine $i$ independently
%\State ~~~~~constructs $\hat W$
%\State ~~~~~randomly selects a dataset $Z^{reopt}_i\subset Z_i$
%\State ~~~~~calculates $\trans\x\hat W$
%\State All machines broadcast $Z^{reopt}_i$ to a master machine
%\State The master calculates $\wreopt$
%\end{algorithmic}
%%\label{fig:alg}
%\end{algorithm}

@@

\cite{liu2014distributed} propose a more Bayesian approach inspired by \cite{merugu2003privacy}.
Instead of averaging the model's parameters,
they directly ``average the models'' with the following KL-average estimator.
\begin{equation}
\wkl = \argmin_\w \sum_{i=1}^m \kl[\bigg]{p(\cdot;\wmle_i)}{p(\cdot;\w)}
\end{equation}
The minimization is performed via a bootstrap sample from the smaller models.
This method has two main advantages.
First, it is robust to reparameterizations of the model.
Second, it is statistically optimal for the class of non-interactive optimization methods.
(We show in the next section that their bound does not apply to our $\wreopt$ estimator due to our semi-interactive setting.)
The main downside of the KL-average is that the minimization has a prohibitively high computational cost.
Let $n^{kl}$ be the size of the bootstrap sample.
Then the original implementation's MSE shrinks as $O((nm)^{-1}+(nn^{kl})^{-1})$.
This implies that the bootstrap procedure requires as many samples as the original problem to get a MSE that shrinks at the same rate as the averaging estimator.
\cite{han2016bootstrap} show a method to reduce this rate to $O((nm)^{-1}+(n^2n^{kl})^{-1})$ using control variates, but the procedure remains prohibitively expensive.
Their experiments show the procedure scaling only to datasets of size $nm\approx10^4$,
whereas our experiments involve a dataset of size $nm\approx10^8$

@@

\subsection {Subgaussian}

%We say that a random vector $X$ is \emph{subgaussian} with mean $\mu$ and covariance $\Sigma$
%if there exists a constant $c>0$ such that
%\begin{equation}
%\prob{\ltwo{\Sigma(X-\mu)} > t} \le c \exp(-t^2)
%\end{equation}
%The simplest example of a subgaussian random vector is a gaussian random vector.
%We denote this relationship in symbols as $X \lesssim \normal{\mu}{\Sigma}$.

We say that a random vector $\x\in\mathbb{R}^d$ is \emph{subgaussian} with variance proxy $\sigma>0$ if for all $\alpha\in\mathbb{R}^d$,
\begin{equation}
\E\exp(\trans\alpha\x) \le \exp(\ltwo{\alpha}^2\sigma^2/2)
.
\end{equation}
The simplest example of a subgaussian random variable is a multivariate normal vector with mean $\zero$ and covariance $I$.
In this case, $\sigma=1$.

\subsection{The Central Limit Theorem}

For completeness,
we state in this section a simplified version of Theorem 5.2 in \cite{spokoiny2012parametricestimation}.
We begin by stating the theorem's assumptions.
These assumptions are general.
We note in particular that logistic regression satisfies these assumptions,
and that these assumptions do not require the likelihood function be convex.

%\begin{enumerate}
%\item
\begin{assumption}
The likelihood function $f(y,\trans\x\w)$ is two times differentiable with respect to $\w$.
\end{assumption}
%\item
\begin{assumption}
For all $u>0$, there exists a function $g_1(u)>0$ such that for all $|\lambda| \le g_1(u)$
and for all $\x\in\X$,
\begin{equation}
\sup_{\w:\ltwo{\I_\wstar(\w-\wstar)} \le u}
\log\E\exp\left(\lambda\frac{\trans\x\nabla_\w f(y_i,\trans\x\w)}{\ltwo{V\x}}\right)
\le
\frac{v_0^2\lambda^2}{2}
\end{equation}
where $\trans V V = \I_\wstar$ and $v_0$ is a constant.
\end{assumption}
%\item
\begin{assumption}
For all $u>0$, there exists a function $b(u)>0$ such that
%\begin{equation}
%\sup_{\w:\ltwo{\I_\wstar(\w-\wstar)} = u}
%\log\E\exp\left(\frac{\E\left(f(y,\trans\x\w) - f(y,\trans\x\wstar)\right)}{\ltwo{\I_\wstar(\w-\wstar)}}\right)
%\le
%b(u)
%\end{equation}
\begin{equation}
\sup_{\w:\ltwo{\I_\wstar(\w-\wstar)} = u}
\log\E\exp\left(\frac{k(\w,\wstar)}{\ltwo{V(\w-\wstar)}}\right)
\le
b(u)
\end{equation}
where
%\begin{equation}
$
k(\w,\wstar) = \E\left(f(y,\trans\x\w) - f(y,\trans\x\wstar)\right)
$
.
%\end{equation}
\end{assumption}
%\end{enumerate}

\begin{theorem}
Assume Conditions 1-3 hold.
Then with probability at least $1-\delta$,
\begin{equation}
\ltwo{V(\w-\wstar)} \le \sqrt{\frac{d+\log{1/\delta}}{n}}
\end{equation}
\end{theorem}

\subsection{Few data points are needed}
\label{sec:vcdim}

Define the space
$\Wave=\affspan\{\wave_i\}_{i=1}^m\cup\{\wave\}$.
Define $L : \mathbb{R} \to \mathbb{R}$ to be the 0-1 loss over the true data distribution.

\begin{theorem}
Let $\w^\Wave$ be an arbitrary point in $\Wave$ and $\delta\in(0,1)$.
Then with probability at least $1-\delta$,
\begin{equation}
L(\wreopt) \le L(\w^\Wave) + O\left(\sqrt{\frac{m + \log(1/\delta)}{n^{reopt}}}\right)
\end{equation}
\end{theorem}
\begin{proof}
For each machine $i$, fix the set of data points $Z_i$.
This in turn fixes the $\wmle_i$ and the space $\Wave$.
The hypothesis class defined by $\Wave$ has VC dimension $m$ because it is linear with $m$ parameters.
Since each of the data points in $Z^{reopt}$ is independent of $\Wave$,
the result follows immediately from standard VC theory.
See for example Chapter 6 of \cite{shalev2014understanding}.
%
%\fixme{
    %Need to rework the definitions in Section 1 so that the data points in the second optimization don't depend on the first optimization.
%}
\end{proof}

\subsection{In the Normal Regime}
%\label{sec:normal}
%
%\begin{theorem}
%We have that with probability at least
%\begin{equation}
%\wmle-\wstar \lesssim \frac{1}{\sqrt{n}}\normal{\zero}{\I^{-1}_{\wstar}}
%\end{equation}
%In the limit as $n\to\infty$, we have the stronger statement that
%\begin{equation}
%\wmle-\wstar \sim \frac{1}{\sqrt{n}}\normal{\zero}{\I^{-1}_{\wstar}}
%\end{equation}
%\end{theorem}

\subsubsection{Assumptions}

We assume that for every estimator $\hat\w$ trained on $n$ data points,
\begin{equation}
\label{eq:assumption}
\hat\w - \E\hat\w \sim \frac{1}{\sqrt n}\normal{\zero}{\I^{-1}_{\E\hat\w}}
%\hat\w \sim \normal{\E\hat\w}{\frac{1}{n}I^{-1}_{\E\hat\w}}
\end{equation}
%
%Throughout Section \ref{sec:normal} we assume that $\wave_i$ is normally distributed with mean $\wtave$ and covariance $\frac{1}{n}I^{-1}_{\wtave}$.
%\begin{equation}
%\label{eq:assumption}
%\wmle_i \sim \normal{\E\wmle_i}{\frac{1}{n}I^{-1}_{\E\wmle_i}}
%\end{equation}
%Here, $I_{\wtave}$ is the Fisher information matrix at point $\wtave$.
Here, $\I_{\E\hat\w}$ is the Fisher information matrix for parameter $\E\hat\w$.
This assumption is motivated by the central limit theorem for maximum likelihood estimation.
%There are many versions of this theorem.
Theorem 7.5.2 of \cite{lehmann1999elements} is particularly simple version and reproduced below.
\begin{theorem}
\label{thm:clt}
Assume the following mild regularity conditions:
\begin{enumerate}[noitemsep,topsep=0pt]%,label=(\itshape\roman*)]
\item all data points in $Z_i$ are i.i.d.;
\item the parameter space $\Theta$ is open;
\item the model is identifiable and consistent;
\item the matrix $\I_{\wtave}$ is non-singular;
\item the support set $A = \{(y,\x) : f(y,\trans\x\theta) > 0\}$ is independent of $\theta$;
and \item for all $(y,\x)\in A$, $f(y,\trans\x\theta)$ is three times differentiable with respect to $\theta$,
and the third derivative is continuous.
\end{enumerate}
Then in the limit as $n\to\infty$,
\begin{equation*}
\wave_i \law \wtave + \frac{1}{\sqrt n} \normal{0}{\I^{-1}_\wstar}
\end{equation*}
Here the notation $\law$ denotes convergence in law,
which is the standard notion of convergence for a sequence of probability distributions.
\end{theorem}

Note that Theorem \ref{thm:clt} does not require the likelihood $f$ to be convex.

There are many generalizations of Theorem \ref{thm:clt}.
For example, Theorem 4.2 in \cite{spokoiny2012parametricestimation} provides finite sample bounds with explicit dependence on the degree of model misspecification, optimization error, and the lack of independence in the data variables.
Because Spokoiny's theorem is non-asymptotic,
the distribution of the estimator is a sub-Gaussian random variable.
This more general setting requires more extensive notation,
but all of our results hold.

\subsubsection{Analyzing $\wave$}

We provide a simple bound that shows that averaging does not improve the bias of our estimator,
but reduces the variance at the rate $O((nm)^{-1})$.
Similar bounds are well known (see Section \ref{sec:alt}).
Our analysis has the following advantages:
it requires fewer assumptions, has a simpler proof, and has an easy to interpret constant factor.
The main idea of our technique is to show that the variance is normally distributed.
The ideas in this proof will be used in Section \ref{sec:analreopt} where we show that the variance of $\wreopt$ is the same as $\wave$, but that the bias is reduced.

\ignore{
\begin{lemma}
We have that
\begin{equation}
\E\ltwo{\wstar-\wave}^2
\le
\E\ltwo{\wstar-\E\wmle_i}^2
+
\frac{1}{m}\E\ltwo{\E\wmle_i-\wmle_i}^2
\end{equation}
\end{lemma}
\begin{proof}
It is well known that the MSE of an estimator is the sum of its bias and variance.
That is,
\begin{equation}
\E\ltwo{\wstar-\wave}^2
=
\E\ltwo{\wstar-\E\wave}^2
+
\E\ltwo{\E\wave-\wave}^2
\end{equation}
First we consider the bias.
By the linearity of expectation, we have that
\begin{align}
\E\wave
&=
\E\frac{1}{m}\sum_{i=1}^m\wmle_i
%\\&=
=
\frac{1}{m}\sum_{i=1}^m\E\wmle_i
=
\E\wmle_i
\label{eq:expwave}
\end{align}
and so the bias
$
\E\ltwo{\wstar-\E\wave}^2
=
\E\ltwo{\wstar-\E\wmle_i}^2
$.
In words, the bias of the averaging estimator is the same as the bias of the estimator on an individual machine.

>>>>

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\vecspan}{span}
\DeclareMathOperator*{\affspan}{aff}
\DeclareMathOperator*{\subG}{subG}
\DeclareMathOperator*{\tr}{tr}

\newcommand{\zero}{\text{\textbf{0}}}
\newcommand{\Zproj}{Z^{\textit{proj}}}
\newcommand{\Zreopt}{Z^{\textit{owa}}}
\newcommand{\nreopt}{n^{\textit{owa}}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\W}{{\hat \Theta^{\textit{owa}}}}
\newcommand{\Wowa}{{\hat \Theta^{\textit{owa}}}}
\newcommand{\Waff}{\mathcal{\hat A}}
\newcommand{\WaffE}{{\mathcal{\hat A}_\E}}
\newcommand{\Wave}{{\mathcal{\hat W}^{ave}}}
\newcommand{\Wtave}{{\mathcal{W}^{ave,*}}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\theta}
\newcommand{\wkl}{\hat\w^{kl}}
\newcommand{\ahat}{\hat\alpha}
\newcommand{\afull}{\ahat^{\textit{full}}}
\newcommand{\wowa}{\hat\w^{owa}}
\newcommand{\wowafull}{\hat\w^{\textit{owa,full}}}
\newcommand{\wowastar}{\hat\w^{\textit{owa,*}}}
\newcommand{\wave}{\hat\w^{ave}}
\newcommand{\wtave}{\E\hat\w^{ave}}
\newcommand{\waver}{\hat\w^{ave,r}}
\newcommand{\wboot}{\hat\w^{boot}}
\newcommand{\wmle}{\hat\w^{mle}}
\newcommand{\wmler}{\hat\w^{mle,r}}
\newcommand{\wstar}{{\w^{*}}}
\newcommand{\wq}{\hat\w^{q}}
\newcommand{\wqstar}{\hat\w^{q^*}}
\newcommand{\dist}{\mathcal{D}}

@@

\newcommand{\I}{\mathcal I}
%\newcommand{\I^{-1}}{\I^{-1}}
\newcommand{\law}{\ensuremath{\xrightarrow{L}}}
\newcommand{\normal}[2]{\ensuremath{\mathcal{N}\left({{#1}},{{#2}}\right)}}
\newcommand{\subnormal}[1]{\ensuremath{\subG\left({{#1}}\right)}}
\newcommand{\trans}[1]{\ensuremath{{#1}^{\mathsf{T}}}}
\newcommand{\ltwo}[1]{{\lVert {#1} \rVert}}
\newcommand{\ltwobig}[1]{{\left\lVert {#1} \right\rVert}}
\newcommand{\lone}[1]{{\lVert {#1} \rVert}_1}
\newcommand{\lzero}[1]{{\lVert {#1} \rVert}_0}
\newcommand{\proj}[1]{\pi_{{#1}}}
\newcommand{\prob}[1]{\Pr\left[{#1}\right]}

@@

@incollection{vershynin2010introduction,
  title={Introduction to the non-asymptotic analysis of random matrices},
  author={Vershynin, Roman},
  journal={arXiv preprint arXiv:1011.3027},
  chapter=5,
  booktitle={Compressed Sensing, Theory and Applications},
  editor={Y. Eldar and G. Kutyniok},
  year={2012}
}

@@

\begin{abstract}
We present a distributed learning algorithm with low communication complexity.
Our algorithm is an easy to implement modification of the standard averaging method.
In the averaging method, each machine in a cluster first solves the learning problem on a subset of the data;
the resulting parameter vectors are then averaged together.
In our method, we perform a second round of optimization to find the \emph{optimal weighted average} of the parameter vectors.
This second optimization happens in a low dimensional space,
and so it adds negligible computational and communication costs.
%Given $m$ machines (each with $n$ data points of dimension $d$),
%we show that our algorithm reduces bias at a rate of $O(\frac{1-m/d}{n})$.
Whereas the averaging method reduces only the variance of the estimator,
our method reduces both the bias and the variance.
Our analysis is more general than the analysis of similar communication efficient algorithms that also reduce the bias.
Notably, we do not assume (as do all previous analyses) that the likelihood function is convex or that any random quantities are bounded.
In practice,
this bias reduction makes our algorithm more robust to the choice of regularization parameter,
speeding up the model selection process.
\end{abstract}

@@

\emph{Non-interactive} algorithms require only a single round of communication.
They are significantly faster than interactive algorithms
but have worse regret bounds.
Recent work (discussed in Section \ref{sec:bounds}) has shown that no non-interactive algorithm can achieve regret bounds comparable to an interactive one.
The popular large scale learning system Vowpal Wabbit cleverly combines both types of algorithms to achieve good real world performance.
A non-interactive algorithm is run first to provide a good warm start to a subsequent interactive algorithm \citep{vw,agarwal2014reliable}.
%Synonyms for non-interactive algorithms used in the literature include \emph{one-shot distributed learning} and \emph{communication efficient learning}.

In this paper, we propose a \emph{semi-interactive} distributed algorithm called \emph{optimal weighted averaging} (OWA).
Our algorithm performs two rounds of communication,
so it is not subject to the existing regret bounds of non-interactive algorithms.
The amount of information communicated in the second round is small, however,
so our algorithm retains the speed advantages of non-interactive algorithms.

@@

To motivate our estimator, we first describe a simpler estimator that uses the entire dataset for the second round of optimization.
%The idea of our proposed solution is to find the best linear combination of the $\wmle_i$s using a second round of optimization.
Define the matrix $\hat W : \mathbb{R}^{d\times m}$ to have $i$th column equal to $\wmle_i$.
Now consider the estimator
\begin{equation}
\wowafull = \hat W \afull
\end{equation}
where
\begin{equation}
\afull = \argmax_\alpha \sum _{(\x,y)\in Z} f\left(y,\trans\x \hat W \alpha \right)
\end{equation}
%This is the maximum likelihood estimator when we restrict our
Notice that $\wowafull$ is the maximum likelihood estimator when the parameter space $\Theta$ is restricted $\W = \vecspan \{\wmle_i\}_{i=1}^m$.
Figure \ref{fig:contour} shows graphically that no other estimator in $\W$ can have higher empirical likelihood than $\wowafull$.

Calculating the weights $\afull$ directly is infeasible because it requires access to the full dataset.
Fortunately, we don't need to consider all the data points for an accurate estimator.
The parameter space $\W$ is $m$-dimensional.
So intuitively, we only need $O(m)$ data points to solve such a problem to our desired accuracy.
%(This intuition is formalized in Section \ref{sec:anal}.)
This intuition motivates the following estimator.
Let $\Zreopt_i\subset Z_i$ be a set of $\nreopt$ data points uniformly sampled from $Z_i$ without replacement,
and let $\Zreopt$ be the union of the $\Zreopt_i$s.
Then the OWA estimator is defined as
\begin{equation}
\wowa = \hat W \ahat
\end{equation}
where
\begin{equation}
\ahat = \argmax_\alpha \sum _{(\x,y)\in \Zreopt} f\left(y,\trans\x \hat W \alpha \right)
\end{equation}

Algorithm \ref{fig:alg2} shows the steps for efficiently computing $\wowa$.
In the first round, each machine calculates $\wmle_i$ independently and broadcasts the results to every other machine.
Since the parameter vector has $d$ dimensions and there are $m$ machines, a total of $O(md)$ bits of get transmitted.
In the second round, each machine projects its local dataset $\Zreopt$ onto the space $\W$.
These projected data points are then transmitted to a predesignated master machine.
The projected data points each have dimension $m$, so a total of $O(m^2\nreopt)$ bits are transmitted.
The master machine now has all of the information to complete the optimization.

@@

\node[dot] at (-1.35,2.03) {};
\node at (-0.9,2.4) {$\wowafull$};

@@

\end{tikzpicture}
\caption{
    Our method performs a second round of optimization to find the best parameter vector in $\W$.
    Since $\wave\in\W$, the estimator $\wowafull$ has higher empirical likelihood than $\wave$ almost surely.
}
\label{fig:contour}
\end{figure}

\begin{algorithm}[t]
\caption{Estimation in two rounds}
\label{alg:distributed}
\begin{algorithmic}
\State Each machine $i$ independently
\State ~~~~~loads dataset $Z_i$
\State ~~~~~calculates $\wmle_i$
\State All machines broadcast $\wmle_i$ to all other machines
\State Each machine $i$ independently
\State ~~~~~constructs $\hat W$
\State ~~~~~randomly selects a dataset $Z^{reopt}_i\subset Z_i$
\State ~~~~~calculates $\Zproj_i=\{(\trans\x\hat W,y) : (\x,y)\in\Zreopt\}$
\State All machines broadcast $\Zproj_i$ to a master machine
\State The master calculates $\wowa$
\end{algorithmic}
\label{fig:alg2}
\end{algorithm}

%\begin{algorithm}[t]
%\caption{Estimation with common data}
%\label{alg:distributed}
%\begin{algorithmic}
%\State Each machine $i$ independently
%\State ~~~~~loads dataset $Z_i$
%\State ~~~~~calculates $\wmle_i$
%\State All machines broadcast $\wmle_i$ to all other machines
%\State Each machine $i$ independently
%\State ~~~~~constructs $\hat W$
%\State ~~~~~randomly selects a dataset $Z^{reopt}_i\subset Z_i$
%\State ~~~~~calculates $\trans\x\hat W$
%\State All machines broadcast $Z^{reopt}_i$ to a master machine
%\State The master calculates $\wowa$
%\end{algorithmic}
%%\label{fig:alg}
%\end{algorithm}

@@

\cite{liu2014distributed} propose a more Bayesian approach inspired by \cite{merugu2003privacy}.
Instead of averaging the model's parameters,
they directly ``average the models'' with the following KL-average estimator.
\begin{equation}
\wkl = \argmin_\w \sum_{i=1}^m \kl[\bigg]{p(\cdot;\wmle_i)}{p(\cdot;\w)}
\end{equation}
The minimization is performed via a bootstrap sample from the smaller models.
This method has two main advantages.
First, it is robust to reparameterizations of the model.
Second, it is statistically optimal for the class of non-interactive optimization methods.
(We show in the next section that their bound does not apply to our $\wowa$ estimator due to our semi-interactive setting.)
The main downside of the KL-average is that the minimization has a prohibitively high computational cost.
Let $n^{kl}$ be the size of the bootstrap sample.
Then the original implementation's MSE shrinks as $O((nm)^{-1}+(nn^{kl})^{-1})$.
This implies that the bootstrap procedure requires as many samples as the original problem to get a MSE that shrinks at the same rate as the averaging estimator.
\cite{han2016bootstrap} show a method to reduce this rate to $O((nm)^{-1}+(n^2n^{kl})^{-1})$ using control variates, but the procedure remains prohibitively expensive.
Their experiments show the procedure scaling only to datasets of size $nm\approx10^4$,
whereas our experiments involve a dataset of size $nm\approx10^8$

@@

We provide a simple bound that shows that averaging does not improve the bias of our estimator,
but reduces the variance at the rate $O((nm)^{-1})$.
Similar bounds are well known (see Section \ref{sec:alt}).
Our analysis has the following advantages:
it requires fewer assumptions, has a simpler proof, and has an easy to interpret constant factor.
The main idea of our technique is to show that the variance is normally distributed.
The ideas in this proof will be used in Section \ref{sec:analreopt} where we show that the variance of $\wowa$ is the same as $\wave$, but that the bias is reduced.
\begin{theorem}
Let $t>0$.
Then with probability at least $1-\exp(-t)$,
\begin{equation}
\ltwo{\wstar-\wave} \le \ltwo{\wstar-\E\wmle_i} + \sqrt\frac{v_t}{mn}
\end{equation}
where
\begin{equation}
v_t =
\tr{\I^{-1}_{\wtave}}
+ 2\sqrt{\tr \left({\I^{-2}_{\wtave}}\right)t}
+ 2\ltwo{\I^{-1}_{\wtave}}t
\end{equation}
\end{theorem}

\subsection {Non-Asymptotic Central Limit Theorem}

The standard central limit theorem for maximum likelihood estimation states that as the number of samples $n \to \infty$,
%the distribution of the parameter vector approaches a Gaussian at the rate $O(n^{-1/2}$).
\begin{equation}
\I^{-1/2}_\wstar(\w-\wstar) \to \frac{1}{\sqrt n}\normal{\zero}{I}
.
\end{equation}
See chapter 7 of \cite{lehmann1999elements} for an elementary presentation of this result.
In the non-asymptotic case, the distribution of the parameter vector is no longer Gaussian;
instead, it is subgaussian.
We say a random vector $\x$ with zero mean is \emph{subgaussian} with variance proxy $\sigma^2$ if
\begin{equation}
\prob{\ltwo{\x} > t} \le \exp(-\sigma^2 t^2 /2)
.
\end{equation}
In notation, we say
\begin{equation}
\x \sim \subnormal{\sigma^2}
.
\end{equation}
%We say that a random vector $\x\in\mathbb{R}^d$ is \emph{subgaussian} with variance proxy $\sigma>0$ if for all $\alpha\in\mathbb{R}^d$,
%\begin{equation}
%\E\exp(\trans\alpha\x) \le \exp(\ltwo{\alpha}^2\sigma^2/2)
%.
%\end{equation}
The simplest example of a subgaussian random variable is a multivariate normal vector with mean $\zero$ and covariance $I$.
In this case, $\sigma=1$.
See \cite{vershynin2010introduction} for a tutorial on subgaussian random variables,
including equivalent definitions and important results.

Before stating the theorem, we first state the theorem's three conditions.
The first condition is straightforward,
but next two conditions are more technical.
Despite the technical appearance of these assumptions,
they are highly general.
We note in particular that any generalized linear model (e.g. least squares, Poisson, and logistic regression),
and many models with nonconvex likelihood functions satisfy these conditions.
We include the conditions here for completeness,
but defer to \cite{spokoiny2012parametricestimation} for a more detailed discussion.

\begin{assumption}
The likelihood function $f(y,\trans\x\w)$ is two times differentiable with respect to $\w$.
\end{assumption}
\begin{assumption}
For all $u>0$, there exists a function $g_1(u)>0$ such that for all $|\lambda| \le g_1(u)$
and for all $\x\in\X$,
\begin{equation}
\sup_{\w:\ltwo{\I_\wstar(\w-\wstar)} \le u}
\log\E\exp\left(\lambda\frac{\trans\x\nabla_\w f(y_i,\trans\x\w)}{\ltwo{V\x}}\right)
\le
\frac{v_0^2\lambda^2}{2}
\end{equation}
where $\trans V V = \I_\wstar$ and $v_0$ is a constant.
\end{assumption}
\begin{assumption}
For all $u>0$, there exists a function $b(u)>0$ such that
\begin{equation}
\sup_{\w:\ltwo{\I_\wstar(\w-\wstar)} = u}
\log\E\exp\left(\frac{k(\w,\wstar)}{\ltwo{V(\w-\wstar)}}\right)
\le
b(u)
\end{equation}
where
$
k(\w,\wstar) = \E\left(f(y,\trans\x\w) - f(y,\trans\x\wstar)\right)
$
.
\end{assumption}

The following is a restatement of Theorem 5.2 in \cite{spokoiny2012parametricestimation}.
\begin{theorem}
Assume Conditions 1-3 hold.
Then whenever $t \le c\sqrt{p/n}$,
we have with probability at least
%Then with probability at least $1-\delta$,
%\begin{equation}
%\ltwo{V(\w-\wstar)} \le c\sqrt{\frac{d+\log{1/\delta}}{n}}
%\end{equation}
%where the constant $c$ depends on $v_o$, $g_1(u)$, and $b(u)$ defined above.
%Equivalently, we have that
%\begin{equation}
%V(\w-\wstar) \sim \subnormal{\sigma}
%\end{equation}
\end{theorem}
